Investigating social change during cultural contact period using geometric morphometry of pottery shapes from Iron Age northeastern Taiwan

### 2020-01-27 analysis/paper/paper.Rmd
starting to edit the discussion

Previous investigations at Kiwulan suggest an unequal distribution pattern of prestige goods, trade ornaments specifically, after the presence of the European that hints the emergence of social inequality. We examine the degree of the ceramic standardization to measure ceramic specialization that is usually associate with social differentiation, such as the elite control [@Costin2001; @Junker1999]. The results confirm that the differences in pottery shapes can be detected using EFA method, indicating the foreign influences on the indigenous society- the Europeans in the 17th century and the Chinese in the 19th century. The result of MANOVA shows there is a significant difference in shapes between pre-European and Post-European period, and between the pre-European and the Chinese period. The average shape presents a round body with a wider rim and neck before the European contact, and turns to oval-shaped body with narrower rim and neck after the European presence and even more obvious during the Chinese contact period. In general, pottery shapes become shorter in hight over time that leads to the oval-shaped body. For the standardization of pottery shape, the result of CV test indicates the significant differences between pre-European period and post-European period (P-value less than 0.05), and between Chinese contact and either pre-European or post-European (P-value less than 0.01). This suggests pottery shape becomes more homogeneous and standardized after foreign contacts with the European colonizer and even more with the Chinese immigrants. The compositional analysis shows clay pastes are similar, regardless of the increasing standardization of the pottery shape, reflecting the raw material sources remains unchanged. More standardized shape in the Chinese period might indicate the appearance of a specialized group. To figure out whether this change is related to the function of pot, we made an attempt to conduct lipid analysis for potsherds samples to identify the source of residue absorbed into the fabric. However, we did not obtain useful results due to a low lipid yield, which might result from the thin and less porous fabric that offers limited spaces to protect organic molecules from microbiological degradation [@Evershed2008, pp. 909].

>>>>

Previous investigations at Kiwulan have suggested an unequal distribution of prestige goods, trade ornaments specifically, following the appearance of Europeans <!-- citation to that previous investigation -->, hinting at the emergence of social inequality in the Indigenous community. Here we have examined ceramic vessel shape standardization to measure craft specialization as a proxy for social differentiation, such as elite control [@Costin2001; @Junker1999]. The results confirm that the differences in pottery shapes can be detected using EFA method, indicating the foreign influences on the indigenous society- the Europeans in the 17th century and the Chinese in the 19th century. The result of MANOVA shows there is a significant difference in shapes between pre-European and Post-European period, and between the pre-European and the Chinese period. The average shape presents a round body with a wider rim and neck before the European contact, and turns to oval-shaped body with narrower rim and neck after the European presence and even more obvious during the Chinese contact period. In general, pottery shapes become shorter in hight over time that leads to the oval-shaped body. For the standardization of pottery shape, the result of CV test indicates the significant differences between pre-European period and post-European period (P-value less than 0.05), and between Chinese contact and either pre-European or post-European (P-value less than 0.01). This suggests pottery shape becomes more homogeneous and standardized after foreign contacts with the European colonizer and even more with the Chinese immigrants. The compositional analysis shows clay pastes are similar, regardless of the increasing standardization of the pottery shape, reflecting the raw material sources remains unchanged. More standardized shape in the Chinese period might indicate the appearance of a specialized group. To figure out whether this change is related to the function of pot, we made an attempt to conduct lipid analysis for potsherds samples to identify the source of residue absorbed into the fabric. However, we did not obtain useful results due to a low lipid yield, which might result from the thin and less porous fabric that offers limited spaces to protect organic molecules from microbiological degradation [@Evershed2008, pp. 909].
